Explore the factors that may be associated with obesity

### 2024-03-23 R/analysis-code/exploratory-analysis.qmd
change dataset and finish part 3

---
title: "An example exploratory analysis script"
author: "Andreas Handel"
date: "10/28/2021"
output: html_document
---

@@

# Data exploration through tables

Showing a bit of code to produce and save a summary table.

```{r}
summary_df = skimr::skim(mydata)
print(summary_df)
# save to file
summarytable_file = here("results", "tables", "summarytable.rds")
saveRDS(summary_df, file = summarytable_file)
```

linear model applied

```{r}
# Fit the ANOVA model
model <- aov(Selective ~ Seed + Strain + Treatment + Day + Rep, data = mydata)

# Summarize the ANOVA results
linear_model <- summary(model)

# Save this table

anovatable_file = here("results", "tables", "anovatable.rds")
saveRDS(linear_model, file = anovatable_file)
```

# Notes

>>>>

---
title: "An example exploratory analysis script"
author: "Xueyan Hu"
date: "03/22/2024"
output: html_document
---

@@

# fit simple linear model

First, I am curious that if physical activity could be correlated to the choice of transportation since I have heard a theory that if someone take much effort on commute transportation, they tend to spend less time on exercise.

```{r}
# Fit multinomial logistic regression model
#library(nnet)
#multinom_model <- multinom(FAF ~ MTRANS, data = mydata)

# Summary of the model
#summary(multinom_model)
```
I will try to figure it out later how I could make it fitting.

Then, I would like to see how drinking water plus physical activity affect obesity.

```{r}
# linear model of BMI by water and FAF
lm1 <- lm(BMI ~ Water + FAF + Water * FAF, data = mydata)
lm1_result <- summary(lm1)

# save the result
lm_file = here("results", "tables", "lm1table.rds")
saveRDS(lm1_result, file = lm_file)
```

In addition, I want to fit model for alcohol consumption and obesity level.

```{r}
mydata$Alcohol <- as.factor(mydata$Alcohol)
lm2 <- lm(BMI ~ Alcohol, data = mydata)
lm2_result <- summary(lm2)

# save the result
lm_file = here("results", "tables", "lm2table.rds")
saveRDS(lm2_result, file = lm_file)
```
I think it makes a little bit sense to me that there is a connection effect of water and physical activity on BMI, which might be supportive for what we learn that drinking more water and doing more exercise will lower the risk of obesity.

Last, I will try to use all predictors that I believe may affect BMI to fit a model.

```{r}
mydata1 <- mydata %>%
  select(History, Water, Alcohol, FAF, MTRANS, BMI)
lm3 <- lm(BMI ~ ., data = mydata1)
lm3_result <- summary(lm3)

# save the result
lm_file = here("results", "tables", "lm3table.rds")
saveRDS(lm3_result, file = lm_file)
```
# Notes

### 2024-03-23 R/eda-code/eda-v2.qmd
change dataset and finish part 3

# Data exploration through tables

Showing a bit of code to produce and save a summary table.

```{r,table1}
```

@@

@@

Seed type first: bar chart

```{r,seed}
```

Then goes strain type: box plot

```{r,strain}
```

Now the effect of the treatment: bar chart with color gradient as an indicator

```{r,treatment}
```

The growth trend of the bacterial population over time: scatterplot with line

```{r,day}
```

linear model applied

```{r,modelfit}
```

# Notes

>>>>

Figures of population distribution by gender and age.

```{r,p1 to p3}
```

@@

Relationship between obesity level and gender/age.

```{r,p4 and p5}
```

Violin plot of obesity level by family history.

```{r,p6}
```

@@

Summary tables.

```{r,table}
```

# note

### 2024-03-23 R/eda-code/eda.qmd
change dataset and finish part 3

```{r}
#load needed packages. make sure they are installed.
library(here) #for data loading/saving
library(dplyr)
library(skimr)
library(ggplot2)
library(maps)
library(gt)
```

@@

```{r}
#Path to data. Note the use of the here() package and not absolute paths
data_location <- here::here("data","processed-data","processeddata_cancer.rds")
#load data
mydata <- readRDS(data_location)
```

# bar chart for crude prevalence according to states

```{r}
# Group by state and calculate the total population and cancer cases
state_total <- mydata %>%
  group_by(StateDesc) %>%
  summarize(TotalPopulation = sum(PopulationCount),
            TotalCancerCases = sum(Cancer_Case))

# Calculate the prevalence for each state
state_total <- state_total %>%
  mutate(TotalPrevalence = TotalCancerCases / TotalPopulation)

# View the resulting dataframe
print(state_total)

# Load usmap package and get the state data
library(usmap)

usmap_state_data <- us_map(regions = "states")

# There are a few state names that are cut off in state_total, but the column
# is already alphabetical so we can replace it.
# To use plot_usmap() we have to have a column named `state` or `fips`.
state_total$state <- usmap_state_data$full
print(state_total)

# Now we can try to make a usmap plot
plot_usmap("states", data = state_total, values = "TotalPrevalence") +
  scale_fill_viridis_c()
```

# List 10 cities that has the highest cancer prevalence

```{r}
top_10_highest <- mydata %>%
  top_n(10, Data_Value) %>%
  select(Year, StateAbbr, CityName, Data_Value)

print(top_10_highest)

# create a table
table <- top_10_highest %>%
  gt() %>%
  tab_spanner(
    label = md('**Year**'),
    columns = c('Year')
  ) %>%
  tab_spanner(
    label = md('**State**'),
    columns = c('StateAbbr')
  ) %>%
  tab_spanner(
    label =  md('**Cancer Prevalence**'),
    columns = c('Data_Value')
  ) %>%
  tab_header(
    title = '2017 top 10 cities which have highest cancer rates'
  ) %>%
  opt_stylize(style = 6, color = 'gray')
```

@@

# fit a simple linear model

Before we started analysis, we had an assumption

```{r}
# Fit linear regression model
lm1 <- lm(Data_Value ~ PopulationCount, data = mydata)

# Print summary of the model
summary(lm1)
```
So the hypothesis is not supported by the linear model fitting result. There is no obvious evidence that the cancer prevalence in larger cities (higher population)

>>>>

```{r}
#load needed packages. make sure they are installed.
library(here) #for data loading/saving
library(dplyr)
library(skimr)
library(ggplot2)
```

@@

```{r}
#Path to data. Note the use of the here() package and not absolute paths
data_location <- here::here("data","processed-data","processeddata.rds")
#load data
mydata <- readRDS(data_location)
```

# gender and age distribution for the cohort under investigation

```{r}
# Convert numeric values to labels
mydata$Gender <- ifelse(mydata$Gender == 1, "Male", "Female")

# bar chart of gender distribution
bc1 <- ggplot(mydata, aes(x = Gender)) +
  geom_bar(fill = c("pink", "skyblue"), color = "black", width = 0.5) +
  geom_text(stat = 'count', aes(label=..count..), vjust = -0.5, size=3, color="black") +
  labs(x = "Gender", y = "Count", title = "Bar Chart of Gender Distribution") +
  theme_minimal()
plot(bc1)

# histogram of population by age
hp2 <- ggplot(mydata, aes(x = Age)) +
  geom_histogram(binwidth = 5, fill = "grey", color = "black") +
  labs(x = "Age", y = "Count", title = "Histogram of Age distribution") +
  stat_bin(binwidth = 5, geom = "text", aes(label = ..count..),
           vjust = -0.5, color = "black", size = 3)  +
  theme_minimal()
plot(hp2)
 # bar chart of obesity level distribution
bc3 <- ggplot(mydata, aes(x = Obesity)) +
  geom_bar() +
  geom_text(stat = 'count', aes(label=..count..), vjust = -0.5, size=3, color="black") +
  scale_x_discrete(labels = c("Under", "Normal", "OW1", "OW2", "Obese1", "Obese2", "Obese3")) +
  labs(x = "Obesity level", y = "Count", title = "Bar Chart of obesity Distribution") +
  theme_minimal()
plot(bc3)
```

# Explore a little bit deeper for Gender and Age

```{r}
unique(mydata$Obesity)
# obesity level by gender
sbp4 <- ggplot(mydata, aes(x = Obesity, fill = Gender)) +
  geom_bar(position = "stack") +
  labs(x = "Obesity Level", y = "Count", title = "Gender Distribution by Obesity Level") +
  scale_fill_manual(values = c("skyblue", "pink"), labels = c("Male", "Female")) +
  scale_x_discrete(labels = c("Under", "Normal", "OW1", "OW2", "Obese1", "Obese2", "Obese3")) +
  theme_minimal()
plot(sbp4)

# Obesity level by age
# Categorize age into groups
mydata <- mydata %>%
  mutate(Age_Group = cut(Age, breaks = c(0, 20, 30, 40, 50, 60, 70, Inf),
                         labels = c("0-20", "21-30", "31-40", "41-50", "51-60", "61-70", "71+")))

sbp5 <- ggplot(mydata, aes(x = Obesity, fill = Age_Group)) +
  geom_bar(position = "stack",stat = "count") +
  coord_flip() + # Rotate the plot to make it horizontal
  labs(x = "Obesity level", y = "Count", title = "Stacked Bar Chart of Obesity Level by Age") +
  theme_minimal()
plot(sbp5)
```
I think the population distribution for age is not balanced since 20-30 age groups contains most of the observations so I will not explore the relationship between obesity and age.

# The relationship between family obesity history and obesity level

```{r}
vp6 <-ggplot(mydata, aes(x = History, y = BMI, fill = History)) +
  geom_violin() +  # Add violin geometry
  geom_boxplot(width = 0.1, fill = "white", color = "black") +  # Add boxplot for better visualization
  geom_jitter(aes(color = History), width = 0.2, alpha = 0.5, size = 1) +   # Add jittered scatterplot
  labs(x = "Family History of Obesity", y = "BMI", title = "Violin Plot of BMI by History of Obesity") +
  scale_fill_manual(values = c("yes" = "lightyellow", "no" = "lightgreen")) +  # Define colors
  theme_minimal()
plot(vp6)
```
# summary table

```{r}
summary1 <- summary(mydata)
summary2 <-skim(mydata)
summarytable_file = here("results", "tables", "summarytable1.rds")
saveRDS(summary1, file = summarytable_file)
summarytable_file = here("results", "tables", "summarytable2.rds")
saveRDS(summary2, file = summarytable_file)
```
# Saving all of the figures

@@

```{r}
figure_file = here("results","figures","gender-distribution.png")
ggsave(filename = figure_file, plot=bc1) 
figure_file = here("results","figures","obesity-distribution.png")
ggsave(filename = figure_file, plot=bc3) 
figure_file = here("results","figures","obesity level by geder.png")
ggsave(filename = figure_file, plot=sbp4) 
figure_file = here("results","figures","obesity level by family history.png")
ggsave(filename = figure_file, plot=vp6) 
```

### 2024-03-23 R/processing-code/processingfile-v1.qmd
change dataset and finish part 3

```{r}
data_location <- here::here("data","raw-data","500_Cities__Cancer__excluding_skin_cancer__among_adults_aged___18_years_20240314.csv")
rawdata <- read.csv(data_location)
```

Several ways of looking at the data

@@

First, only crude prevalence is interested since we don't know the age structure here. And the prevalence of each city is the summary of all the census tracts from this city, so City will be selected only for further analysis. There are also 2 summed data including all the US prevalence that will not be used for analysis.

```{r}
d1 <- rawdata %>%
  filter(GeographicLevel == "City", DataValueTypeID == "CrdPrv", StateAbbr != "US")
```

Dropping uninterested variables for a smaller dataset.

```{r}
d2 <- d1 %>%
  select(Year, StateAbbr, StateDesc, CityName, DataValueTypeID, Data_Value,PopulationCount)
```

Generating a new variable in case it will be needed later.

```{r}
d3 <- d2 %>%
  mutate(Cancer_Case = round((Data_Value / 100) * PopulationCount))
```

Re-check with the data

```{r}
dplyr::glimpse(d3)
summary(d3)
head(d3)
skimr::skim(d3)
```

@@

```{r}
processeddata <- d3
save_data_location <- here::here("data","processed-data","processeddata_cancer.rds")
saveRDS(processeddata, file = save_data_location)
```

>>>>

```{r}
data_location <- here::here("data","raw-data","ObesityDataSet_raw_and_data_sinthetic.csv")
rawdata <- read.csv(data_location)
```

#### check on codebook

```{r}
codebook <- here::here("data","raw-data","codebook.xlsx")
read_excel(codebook)
```
Several ways of looking at the data

@@

First generating a new variable BMI using height and weight, as well as change some variable names and categorical levels.

```{r}
d1 <- rawdata %>%
  mutate(BMI = Weight / (Height)^2) %>%
  rename(History = family_history_with_overweight, Water = CH2O, Alcohol = CALC, Obesity = NObeyesdad) %>%
  mutate(Gender = if_else(Gender == "Male", 1, 2))

d2 <- d1 %>% 
  mutate(Alcohol = case_when(
    Alcohol == "no" ~ 1,
    Alcohol == "Sometimes" ~ 2,
    Alcohol == "Frequently" ~ 3,
    Alcohol == "Always" ~ 4,
    TRUE ~ NA_integer_  # Handle any other cases
  ))
# Print the updated dataframe
head(d2)
```

My interest variables are gender, age, alcohol consumption, water drinking, transportation, physical activity, and outcomes will be obesity level and BMI.

```{r}
d3 <- d2 %>%
  select(Gender, Age, History, Water, Alcohol, FAF, MTRANS,BMI, Obesity)
head(d3)
```

@@

```{r}
processeddata <- d3
save_data_location <- here::here("data","processed-data","processeddata.rds")
saveRDS(processeddata, file = save_data_location)
```

### 2024-03-23 products/manuscript/Manuscript copy.qmd
change dataset and finish part 3

---
title: "Examine the efficacy of ascaroside #18 in controling Escherichia Coli on alfalfa and fenugreek seeds and sprouts"
subtitle: ""
author: Xueyan Hu
date: "`r Sys.Date()`"
format:
  docx:
    toc: false
    number-sections: true
    highlight-style: github
bibliography: dataanalysis-references.bib
csl: apa.csl
editor: 
  markdown: 
    wrap: sentence
---

The structure below is one possible setup for a data analysis project (including the course project).
For a manuscript, adjust as needed.
You don't need to have exactly these sections, but the content covering those sections should be addressed.

This uses MS Word as output format.
[See here](https://quarto.org/docs/output-formats/ms-word.html) for more information.
You can switch to other formats, like html or pdf.
See [the Quarto documentation](https://quarto.org/) for other formats.

@@

In this project, I will be concentrating on analyzing my own data gotten from the experiment that I carried on by my own.
And I will be trying to figure out if there will be any relation between interested variables and observations.

@@

The core idea of this research comes from the following facts.
First, sprouts are usually considered nutritious but they are consumed in raw mainly for foods like salad and sandwiches.
Therefore, there have been a lot of foodborne pathogen outbreaks occurred across the states in the decades [@miyahira2021], which means that there is an urgent need to reduce the contamination of sprouts with foodborne pathogens.

At the meantime, a recently study has revealed that a newly derived chemical could enhence the immune resistance of some certain plants to the pathogens including bacteria and virus but not by killing them directly[@manosalva2015].

Based on the facts above, a hypothesis is raised that if this chemical can also be applied on seedling seeds to control the level of contamination during sprouting process.

@@

The data is obtained from experiment that I did it by myself, which has 160 observations and several variables.
The data is the result of bacterial populations of foodborne pathogens on sprouts that are collected from different types of plants at different germinating time points under different treatments for the seeds

@@

I would like to figure out the effect of different treatments on controlling bacterial populations as well as the factors also may be related to the effectiveness, including seed type for sprouts, bacterial strain type infecting sprouts, etc.
So the treatment type will be the most important predictor that I will focus on.
And I am willing to explore if there is any treatment that could reduce bacterial population level on seedling sprouts.

@@

Since the dataset is quite well-organized and no any value missing, the cleaning step is relatively easy.
I only changed the category names.

My main interest is to test if the chemical treatment is effective or not compared to the control treated groups.
Also, anova model will be applied to check which variable will make significant different to the bacterial population.
So boxplot, bar chart and line chart will be created according to the analysis direction.

@@

## Data aquisition

All of the data in this dataset was collected by me in 2022 after experiment was conducted.

@@

The detailed raw data and processed data are stored in 'data' folder.

*Write code that reads in the file and cleans it so it's ready for analysis. Since this will be fairly long code for most datasets, it might be a good idea to have it in one or several R scripts. If that is the case, explain here briefly what kind of cleaning/processing you do, and provide more details and well documented code somewhere (e.g. as supplement in a paper). All materials, including files that contain code, should be commented well so everyone can follow along.*

@@

*Explain anything related to your statistical analyses.*

@@

*Use a combination of text/tables/figures to explore and describe your data. Show the most important descriptive results here. Additional ones should go in the supplement. Even more can be in the R and Quarto files that are part of your project.*

@@

*To get some further insight into your data, if reasonable you could compute simple statistics (e.g. simple models with 1 predictor) to look for associations between your outcome(s) and each individual predictor variable. Though note that unless you pre-specified the outcome and main exposure, any "p\<0.05 means statistical significance" interpretation is not valid.*

@@

>>>>

---
title: "Explore the factors that may be associated with obesity"
subtitle: ""
author: Xueyan Hu
date: "`r Sys.Date()`"
format:
  docx:
    toc: false
    number-sections: true
    highlight-style: github
bibliography: dataanalysis-references.bib
csl: apa.csl
editor: 
  markdown: 
    wrap: sentence
---

@@

In this project, I have found......

@@

[Obesity](https://www.who.int/news-room/fact-sheets/detail/obesity-and-overweight)![]() is a chronic complex disease defined by excessive fat deposits that can impair health.
Obesity can lead to increased risk of type 2 diabetes and heart disease, it can affect bone health and reproduction, it increases the risk of certain cancers.

In the recent decades, obesity has become a more and more common and severe problem across the world.

Worldwide adult obesity has more than doubled since 1990, and adolescent obesity has quadrupled.
In 2022, 1 in 8 people in the world were living with obesity.

There is no doubt that obesity is a multifactorial disease due to obesogenic environments, psycho-social factors and genetic variants in most cases, but personal behavior could also has correlation with the result of overweight and obesity.

There is a lot of discussion on the Internet and social media about fitness and weight loss, and people assume that physical activity can lose weight as a fact.
Also KOLs ususally encourage people to drink more water for higher metabolic levels.
But I have not seen a clear data to prove it.
At the same time, the effect of physical activity to lose weight is also different from person to person.

In addition, according to what I learned from nutrition course, alcohol consumption may be related to weight gain because [alcohol metabolism](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC6527027/) and carbohydrate metabolism both go through the same pathway/cycle in the middle and late stages.

@@

The data is obtained from a website [UCI Machine Learning Repository](https://archive.ics.uci.edu/dataset/544/estimation+of+obesity+levels+based+on+eating+habits+and+physical+condition) which have collected the survey results of individuals from the countries of Mexico, Peru and Colombia on their eating habits and physical condition, as well as their estimation of obesity levels.

@@

I would like to figure out if higher frequency of physical activity and drinking more water could keep people away from obesity and whether heredity is an important factor for obesity.
Also, I will focus on the relationship between alcohol consumption frequency and obesity level.

@@

Data cleaning including generating necessary variable, renaming categories and variables and so on.

First to take a deeper look of the data, I would like to do some descriptive analysis including bar chart, boxplot, scatterplot etc. to indicate the distribution of the data.

Then simple linear fit will be applied to explore the relationship between variables and outcomes (both categorical and numeric).

Prediction will be made.

@@

## Data acquisition

The dataset presents the results of a survey on the topic of obesity including personal physical conditions and answers to behavioral questions.

@@

The detailed raw data and processed data are stored in 'data' folder.
And the code for data cleaning is in the 'processing-code' subfolder under 'R' folder.

Obesity level is a categorical variable defined by BMI range, so BMI as a new variable is created and it could be analyzed as a numeric variable.
Also, to make the variable name easy to understand, I changed some abbreviated names to ones with fully spelled key words.
Male is replaced by 1 and female by 2 under Gender variable for more convenient further analysis.

@@

Bar chart, box plot will show the distribution of obesity level by different variables, as well as the population features of the people under investigation.
All figures will be saved in 'figure' subfolder under 'result' folder and shown in the manuscript.

By fitting the data with simple linear model, the result will show whether each independent variable has main effect to the dependent variable through the P value.
And the summary table is stored in 'table' subfolder under 'result' folder.

@@

Bar chart shows that the genders of the cohort under investigation are balanced.

Figure 1.
Gender distribution.

![](images/gender-distribution.png)

It can be seen that almost two thirds of the people are identified with over weight.

Figure 2.
Obesity level distribution.

![](images/obesity-distribution.png)

There is no obvious pattern of obesity levels by gender.

Figure 3.
Gender distribution in each obesity levels.

![](images/obesity%20level%20by%20gender.png)

It is apparent that people with family obesity history may tend to be over weight according to figure 4 below.

Figure 4.
Violin plot of BMI by family obesity history.

![](images/obesity level by family history-02.png)

@@

Since some obvious relationship could be shown in figures, I choose other variables as predictors in exploratory analysis

First, I would like to fit linear model for BMI by both water consumption and physical activity.
And the result indicates that there is a connection effect between them two, which means that both of them contribution to obesity control.

Also, I use single indicator, alcohol consumption and BMI as an outcome to fit linear model.

In the last model I select all the variables that I think may have effect on BMI as predictors and BMI as outcome.

@@

**Haven't updated yet but will do it later!!!**
